Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-10-03 chapters/chapter01.tex
plots, formatting, greyscales

\begin{remark}{Outline}
In this chapter, we introduce fundamental 
TODO
\end{remark}

@@

Combating viral spread and its associated disease burden is a tremendous challenge requiring significant research efforts and decided public health measures. 
Viral sequence data has been recognized as a major asset in the characterization of pathogens that threaten public health.
Molecular epidemiology complements traditional epidemiological studies because it primarily focuses on the etiological agent rather than the host \citep{Leitner2002}.
In particular, the historical information contained in viral gene sequences contributes to a better insight in emergence and early transmission dynamics, even before systematic epidemiological surveillance has been initiated. 
For many years, phylogenetic analysis has been the primary tool in molecular epidemiology. 
By reconstructing a ime-calibrated phylogenetic tree, we are able to study the evolutionary history of viral strains, assess epidemiological linkage amongst them, and shed light on transmission patterns. 
With the advent of high-throughput DNA sequencing, genomic data from RNA viruses is becoming available in unprecedented quantities and with remarkable rapidity.

@@

\subsection{Bayesian evolutionary inference in BEAST}
% [PL: How this all integrates in a flexible framework and publicly available software, one that adequately accounts for phylogenetic uncertainty by averaging over all plausible evolutionary histories for your data
% ]

@@

\subsection{Objectives: model extension, simulation and visualisation}

@@

\begin{figure}[H]
\centering
\includegraphics[scale=0.5]{poisson}
\caption{
{ \footnotesize 
{\bf Independent realizations of the Poisson process with intensity $r=1$.}
Times between mutations (the branch lengths) are \emph{iid} exponentially distributed random variables with mean $1/r$.
} % END: footnotesize
}
\label{fig:poisson}
\end{figure}

>>>>

\begin{remark}{Outline}
% TODO
In this chapter, we introduce fundamental concepts of the molecular evolution and how it integrates into the field of phylodynamics, a study of evolutionary and epidemiological processes exhibited by emerging epidemics.
Beginning with a current state of the field, we exemplify some of those epidemics, an illustration of the vulnerability of the human population to pandemic threats.
We stress the importance of developing fast and statistically sound surveilance methods, serving as a a shield against those threats.

We then introduce some of the most important breakthroughs in the field, i.e. the introduction of molecular clock, which allows to date speciation events from genetic data and calibration points alone, and the shift of the field towards probabilistic methods and the advent of Monte Carlo algorithms as well as powerful and increasingly more attainable computer hardware, which boosted the adoption of Bayesian methodology.
We then talk about how these advances integrate into publicly available software, that adequately accounts for phylogenetic uncertainty by averaging over all plausible evolutionary histories.

Finally Subsection~\ref{sub:objectives} positions the work presented in this thesis and how it advances our understanding of the processes shaping viral evolution and spread.
Next parts of this introductory chapter are devoted towards describing the most basic inference methodology in more details, and serve as an aid in understanding the work presented in Chapters~\ref{chap:epoch}--\ref{chap:pibuss}.

\end{remark}

@@

Combating viral spread and its associated disease burden is a tremendous challenge requiring significant research efforts and decided public health measures. 
Viral sequence data has been recognized as a major asset in the characterization of pathogens that threaten public health.
Molecular epidemiology complements traditional epidemiological studies because it primarily focuses on the etiological agent rather than the host \citep{Leitner2002}.
In particular, the historical information contained in viral gene sequences contributes to a better insight in emergence and early transmission dynamics, even before systematic epidemiological surveillance has been initiated. 
For many years, phylogenetic analysis has been the primary tool in molecular epidemiology. 
By reconstructing a time-calibrated phylogenetic tree, we are able to study the evolutionary history of viral strains, assess epidemiological linkage amongst them, and shed light on transmission patterns. 
With the advent of high-throughput DNA sequencing, genomic data from RNA viruses is becoming available in unprecedented quantities and with remarkable rapidity.

@@

\subsection{Bayesian evolutionary inference in BEAST}

@@

\subsection{Objectives: model extension, simulation and visualisation\label{sub:objectives}}

@@

\begin{figure}[H]
\centering
\includegraphics[scale=0.4]{poisson}
\caption{
{ \footnotesize 
{\bf Independent realizations of the Poisson process with intensity $r=1$.}
Times between mutations (the branch lengths) are \emph{iid} exponentially distributed random variables with mean $1/r$.
} % END: footnotesize
}
\label{fig:poisson}
\end{figure}

### 2014-10-03 chapters/chapter06.tex
plots, formatting, greyscales

The GY94-F61 model has been suggested as important in letting the model explain the data without constricting assumptions.
%GB: 'confounds other effects' is too vague, be more specific or give an example
\citet{Rodrigue2008} however argue that the model has no interpretation on the nucleotide level as well as confounds other effects. 
The authors criticize the GY94-F$3\times4$ setup as superficial, pointing that the periodic pattern along the nucleotide sequence positions represents the coding structure of the sequence but not the processes that shape the evolution.

@@

%GB: fix this, it's too wide for the page
\footnotesize{
\begin{equation}
q_{ij}^{MG94}=\begin{cases}
\alpha\cdot\kappa\cdot\pi_{n}^{p} & i\rightarrow j\text{ is a synonymous transition from nucleotide \textit{m} to \ensuremath{n} at codon position \ensuremath{p}.}\\
\alpha\cdot\pi_{n}^{p} & i\rightarrow j\text{ is a synonymous transversion from nucleotide \ensuremath{m} to \ensuremath{n} at codon position \ensuremath{p}.}\\
\beta\cdot\kappa\cdot\pi_{n}^{p} & i\rightarrow j\text{ is a non-synonymous transition from nucleotide \ensuremath{m} to \ensuremath{n} at codon position \ensuremath{p}.}\\
\beta\cdot\pi_{n}^{p} & i\rightarrow j\text{ is a non-synonymous transversion from nucleotide \ensuremath{m} to \ensuremath{n} at codon position \ensuremath{p}.}\\
0 & \text{otherwise.}
\end{cases},
\label{eq:mg94}
\end{equation}
} % END: size

@@

>>>>

The GY94-F61 model has been suggested as important in letting the model explain the data without constricting assumptions.
\citet{Rodrigue2008} however argue that the model has no interpretation on the nucleotide level as well as confounds other effects inducing uneven codon stationary probabilities.
The authors criticize the GY94-F$3\times4$ setup as superficial, pointing that the periodic pattern along the nucleotide sequence positions represents the coding structure of the sequence but not the processes that shape the evolution.

@@

{\tiny{
\begin{equation}
q_{ij}^{MG94}=\begin{cases}
\alpha\cdot\kappa\cdot\pi_{n}^{p} & i\rightarrow j\text{ is a synonymous transition from nucleotide \textit{m} to \ensuremath{n} at codon position \ensuremath{p}}\\
\alpha\cdot\pi_{n}^{p} & \text{synonymous transversion from \ensuremath{m} to \ensuremath{n} at position \ensuremath{p}}\\
\beta\cdot\kappa\cdot\pi_{n}^{p} & \text{non-synonymous transition from \ensuremath{m} to \ensuremath{n} at position \ensuremath{p}}\\
\beta\cdot\pi_{n}^{p} & \text{non-synonymous transversion from \ensuremath{m} to \ensuremath{n} at position \ensuremath{p}}\\
0 & \text{otherwise}
% \alpha\cdot\kappa\cdot\pi_{n}^{p} & i\rightarrow j\text{ is a synonymous transition from nucleotide \textit{m} to \ensuremath{n} at codon position \ensuremath{p}.}\\
% \alpha\cdot\pi_{n}^{p} & i\rightarrow j\text{ is a synonymous transversion from nucleotide \ensuremath{m} to \ensuremath{n} at codon position \ensuremath{p}.}\\
% \beta\cdot\kappa\cdot\pi_{n}^{p} & i\rightarrow j\text{ is a non-synonymous transition from nucleotide \ensuremath{m} to \ensuremath{n} at codon position \ensuremath{p}.}\\
% \beta\cdot\pi_{n}^{p} & i\rightarrow j\text{ is a non-synonymous transversion from nucleotide \ensuremath{m} to \ensuremath{n} at codon position \ensuremath{p}.}\\
% 0 & \text{otherwise.}
\end{cases},
\label{eq:mg94}
\end{equation}
} % END: size
}

@@

%bible p7
